---
title: "Uncap the House! The Case for Increased Representation"
format: html
theme: cosmo
editor: visual
embed_resources: true
code-fold: true
---


```{r}
#| eval: false
install.packages("tidytuesdayR")
```

```{r}
#| output: false
library(tidytuesdayR)
tuesdata <- tidytuesdayR::tt_load('2024-11-05')
library(tidyverse)
```

```{r}
dem_df<- tuesdata$democracy_data
```

```{r}
#| output: false
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
library(readxl)
library(ggrepel)
library(broom)
library(scales)
library(knitr)
library(kableExtra)
```

```{r}
base_url <- "https://api.api-ninjas.com/v1/population"

api <- Sys.getenv("API_NINJAS_KEY")
```


```{r}

get_pop_data <- function(country) {
  country_url <- glue("{base_url}?country={URLencode(country)}")
  countries <- GET(country_url, add_headers("X-Api-Key" = api))
  tx <- content(countries, "text", encoding = "UTF-8")
  if (tx==""){
    return(NA)
  }
  countries_data <- fromJSON(tx)
  if(is.null(countries_data$historical_population)) {
    return(NULL)
  }
  countries_data <- countries_data$historical_population %>%
    mutate(country = country)
  return(countries_data)
}

```

```{r}
#|eval: false
pop_data <- map(unique(dem_df$country_name), .f = get_pop_data)
```

```{r}
#| eval: false
saveRDS(pop_data,"pop_data.rds")
```

```{r}
pop_data <- read_rds("pop_data.rds")
```

```{r}
clean_df <- pop_data %>%
  keep(~ length(.x)>0) %>%
  map(~.x %>% select(country, year,population)) %>%
  bind_rows()
```

```{r}
dem2_df <- dem_df %>%
  filter(year>=1955 & year%%5==0)%>%
  rename(country = "country_name")
```

```{r}
clean2_df <- clean_df %>%
  filter(year%%5==0) %>%
  arrange(country, year)
```

```{r}
final_df <- left_join(dem2_df, clean2_df, by = c("country", "year"))
```

```{r}
final_df <- final_df %>%
  filter(is_democracy==TRUE)%>%
  mutate(pop_prop = population/rowSums(cbind(upper_house_members,lower_house_members, third_house_members), na.rm=TRUE))

# head(final_df[c("country", "year", "population","pop_prop")])
```

```{r}
#| output: false

happiness <- read_csv("whr_data.csv")
# I'm not sure why there are empty columns beyond those we are using, the file very clearly only had 13 columns. I don't think this should be an issue, but we're not gonna be using that whole dataset anyway.
```

```{r}
#| output: false
happiness_clean <- happiness %>%
  select(
    country_name = `Country name`,
    year = Year,
    well_being = `Life evaluation (3-year average)`)

#find the years that both this data set and the dem_df covers, and then we'll filter the sets to match those years

# range(happiness_clean$year, na.rm = TRUE) # since 2011
# range(dem_df$year, na.rm = TRUE) # 1950 to 2020

dem_overlap <- dem_df %>%
  filter(year >= 2011)
happiness_overlap <- happiness_clean %>%
  filter(year <= 2020)

# now we merge them into one data frame
#hdiXdem_df <- dem_overlap %>%
  #left_join(happiness_overlap, by = c("country_name", "year"))


# this will weed out the countries that we weren't able to directly match by name
# hdiXdem_df %>%
# filter(is.na(well_being)) %>%
#  distinct(country_name) %>%
#  arrange(country_name)
# so that gave 208 countries which surprised me because the names looked the same to me. this makes me believe i messed up on the joining of the dataframes.
# i'm gonna try the join again, only based off name, not year as well

# hdiXdem_df <- dem_overlap %>%
 # left_join(happiness_overlap, by = "country_name")

# okay that isn't what we want either, that gives us a many to many dataframe, which is bad, so what I'll do is take the average of the happiness index and democracy index across the years and compare those. For reference, we don't want the many to many dataframe because it has 13010 rows.

# average of both data frames
avg_dem <- dem_overlap %>%
  group_by(country_name) %>%
  summarize(avg_democracy = mean(spatial_democracy, na.rm = TRUE))

avg_happiness <- happiness_overlap %>%
  group_by(country_name) %>%
  summarize(avg_well_being = mean(well_being, na.rm = TRUE))

# new dataframe

hdiXdem_df <- avg_dem %>%
  left_join(avg_happiness, by = "country_name")

# now check again for unmatched countries
hdiXdem_df %>%
  filter(is.na(avg_well_being)) %>%
  distinct(country_name) %>%
  arrange(country_name)
#so that gives us 60 countries that are not in our happiness set that are in the democracy set. What we'll do now is use anti_join to see what countries are in the democracy set that are not in the happiness set, and vice versa.

dem_not_happy <- anti_join(avg_dem, avg_happiness, by = "country_name")
happy_not_dem <- anti_join(avg_happiness, avg_dem, by = "country_name")

# nrow(dem_not_happy)
# nrow(happy_not_dem)

# head(dem_not_happy, 20)
# happy_not_dem

# after looking through it, it seems that the countries that were in dem_not_happy were a lot of small, island based nation states that don't necessarily need to be included for this to be a fair comparison. The countries in happy_not_dem were all countries with names that may be different between the two sets. these are the countries we will recode. Using my own imperfect analysis, I have chosen the countries I will use from this list of unmatched country names, and I will use this to find the names for those countries in the democracy df

avg_dem %>% filter(str_detect(country_name, "Korea|Mold|Congo|Trinidad|Hong|Palest|Taiwan|Turkey|Czech|Gambia|Slov"))

# now that we know the names of all the countries that we want in both data sets, I will fix the happiness dataset to match the names from the democracy set, and then we will use an inner join to only get the countries that we want.

country_fix_happiness <- c(
  "Congo" = "Congo, Republic of",
  "Czechia" = "Czech Republic",
  "Côte d’Ivoire" = "Ivory Coast",
  "DR Congo" = "Congo, Dem. Rep.",
  "Gambia" = "Gambia, The",
  "Hong Kong SAR of China" = "Hong Kong",
  "Lao PDR" = "Laos",
  "Republic of Korea" = "Korea, Republic of",
  "Republic of Moldova" = "Moldova",
  "Russian Federation" = "Russia",
  "Slovakia" = "Slovak Republic",
  "State of Palestine" = "Palestine",
  "Taiwan Province of China" = "Taiwan",
  "Trinidad and Tobago" = "Trinidad &Tobago",
  "Türkiye" = "Turkey",
  "Viet Nam" = "Vietnam")

avg_happiness_fixed <- happiness_overlap %>%
  mutate(country_name = recode(country_name, !!!country_fix_happiness)) %>%
  group_by(country_name) %>%
  summarize(avg_well_being = mean(well_being, na.rm = TRUE), .groups = "drop")

hdiXdem_df <- avg_dem %>%
  inner_join(avg_happiness_fixed, by = "country_name")

# head(hdiXdem_df)
```

## Introduction

Geopolitical structures are very diverse, with everything from parliamentary democracies to theocratic authoritarian regimes. In the United States, the concepts of democracy and representation have basically become normative and almost inherently ethical, so much so that this reasoning has been the basis for heavy involvement and regime change across the world. With this, it is important to take a step back and wonder, does this even matter? Is the democratic structure the most conducive for well-being? Are our structures of representation meaningful when it comes to our qualities of democracy? As norms get disrupted, and streaks of populism keep growing, it is crucial to analyze if our established structures really do bolster well-being for its constituents. 

## Does a country's representation even relate to their democracy quality?

We first use a dataset made avaiable through TidyTuesdayR. It contains information from 1955 to 2020 about the geopolitical structure of every country in the world. Example variables include democracy status, a regime category index, and communist status. The collection of this data has largely been ongoing, but I believe was compiled fairly recently. This kind of data is usually easily attainable because most variables are yes/no questions such as, "Was this country a democracy?". Each observational unit is a combination of country and year, so e.g. Albania 1962. This is a very interesting dataset because with the presence of variables such as that give scores to a country's democracy, hold the number of representatives, and even test if they have free and fair elections, you can make some really interesting conclusions about what we consider as some of the most sacred aspects of a society. For example, an idea that we have is testing a relationship with spatial democracy index vs. total representation to test the long held and intuitive belief that more representation per person leads to better democracies. To obtain this population data, we used an API to collect historical population data. This data was collected every five years since 1955, and is widely accessible through public historical population records. This data was of interest to us because making conclusions about the efficacy of representation is not really useful without some adjusting for population. For example, a country like china may have a lot of representatives but this could be disanalogous with their democracy quality because of the fact that China is the most populated country in the world.


```{r}
democracy_data <- tuesdata$democracy_data
countries <- unique(democracy_data$country_name)
total_reps <- numeric(length(countries))

for (i in seq_along(countries)) {
  this_country <- countries[i]
  
  total_reps[i] <- sum(
    democracy_data$lower_house_members[democracy_data$country_name == this_country],
    na.rm = TRUE)}

country_rep_summary <- data.frame(
  country_name = countries,
  total_lower_house_members = total_reps)

country_rep_summary <- country_rep_summary[order(-country_rep_summary$total_lower_house_members), ]


knitr::kable(
  head(country_rep_summary, 20),
  col.names = c("Country", "Total Lower House Members"),
  caption = "Total lower house members by country")
```

This table highlights countries with the largest total number of lower house representatives. Larger and more populous countries tend to dominate this list, reflecting the scale of their political systems. As we prepare to analyze the representation score per country, we need to gather total numbers of representatives across chambers. This can present itself as a challenge because we have encountered the fact that countries have different ways of organizing their representatives. For example, while the United States has a bicameral legislature, there is an additional column in the dataset for a third house, which we had to account for. Overall, this table both helps us get a good idea for the front runners of representation, and at the same time helps us prepare to gather population data to take some averages to get a population adjusted representation measure.


```{r}
dem_index<-read.csv("https://raw.githubusercontent.com/egregi0us/political-system-analysis/refs/heads/main/democracy-index-eiu.csv")

dem_index2<-dem_index %>%
  rename(country = "Entity")%>%
  rename(year="Year")%>%
  filter(year>2006 & year%%5==0)
```

```{r}
finalest_df <- final_df %>%
  filter(!pop_prop == Inf)
```

```{r}
finalestest_df <- left_join(finalest_df, dem_index2, by = c("country","year"))
```

```{r}
finalestest_df<- finalestest_df %>%
  filter(year>2005)%>%
  filter(!is.na(Democracy.Index))

```

```{r}
finalestest_df %>%
  select(country, year, spatial_democracy, Democracy.Index, population, pop_prop) %>%
  slice_max(order_by = Democracy.Index, n=15) %>%
  mutate(
    spatial_democracy = round(spatial_democracy, 3),
    Democracy.Index = round(Democracy.Index, 3),
    population = scales::comma(population),
    pop_prop = scales::comma(round(pop_prop, 0))
  ) %>%
  knitr::kable(
    col.names = c(
      "Country",
      "Year",
      "Spatial Democracy",
      "Democracy Index",
      "Population",
      "People per Representative"
    ),
    caption = "Democracy, population, and representation summary for selected countries.",
    align = "c"
  )
```

This table shows how population, representation, and democracy scores vary across countries and years. All of these variables are extraordinarily useful for our questions, as it provides context for understanding how representation per person may relate to democratic outcomes. It includes two measures of democracy, both spatial democracy score and democracy index. These were pulled from different sources, and really only differ by the fact that the democracy index seems to be less clustered around certain values between 0 and 1.


```{r}
library(countrycode)
vis2 <- final_df |>
  select(country, spatial_democracy, pop_prop) |>
  mutate(region = countrycode(country,
                              origin = "country.name",
                              destination = "continent")) |>
  filter(!pop_prop==Inf)
```

```{r}
#| output: false
library(cowplot)
library(ggridges)
library(scales)
```

```{r}
#| warning: false
title_theme <- ggdraw() +
  draw_label("Spatial Democracy Versus Person per Representative by Region",
    fontfamily = "Courier",
    fontface = "bold",
    size = 9,
    hjust = 0.42
  )

spa_dem <- ggplot(vis2, aes(x = spatial_democracy,
                            y = reorder(region, spatial_democracy),
                            fill = region)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) + 
  scale_x_continuous(limits = c(0, 1)) +
  labs(x = "Spatial Democracy Index",
       y = " ") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = "none")

pop_dem <- ggplot(vis2, aes(x = pop_prop,
                            y = reorder(region, pop_prop),
                            fill = region)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  scale_x_continuous(limits = c(0, 50000), labels = label_number(scale_cut = cut_short_scale())) +
  labs(x = "Person per Rep",
       y = " ") +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = "none")

gridofplots <- plot_grid(spa_dem, pop_dem, nrow = 1)

plot_grid(title_theme,
  gridofplots,
  ncol = 1, rel_heights = c(0.2, 1.5, 0.1)
)
```

These plots show how both democracy scores and representation levels vary across regions. This is useful to us because we can see how spatial democracy and representation are distributed by continent. Continental patterns can lead to more questions such as what factors that are present continent wide can lead to such disparities? A famous example of these sorts of questions are present in "Guns, Germs, and Steel" by Jared Diamond, where he attributes the disparities between Africa and parts of Asia to the presence of certain crops and livestock, as well as geographical differences.


```{r}
#| warning: false
final_df %>%
  ggplot(aes(x=pop_prop, y = spatial_democracy))+
  geom_jitter(alpha = 0.3)+
  labs(
    x = "Population Proportion",
    y = "Spatial Democracy",
    title = "Jitter Plot, Population Proportion v. Spatial Democracy"
  )
```

This scatterplot shows a wide spread in democracy scores across different levels of population per representative. While there is some visible pattern, the relationship appears noisy and not strictly linear. This is largely due to a shortcoming in our data, being that spatial democracy is a categorical-like quantitative variable, meaning that even though its measured in numbers, its largely easy numbers to work with between 0 and 1, so 0.5,0.6,0.25 for example. This leads to clustering as you can see in the scatterplot. However, past this we can still see some negative trend in the data, indicating a potential negative relationship between spatial democracy and population adjusted representation. Note that the way that we measured representation leads to a higher number meaning lower representation since our population adjusting was measured in person per representative.

## Does the quality of a country's democracy even relate to their overall happiness?

The data pertaining to this research question was read in through a csv file accessed through the World Bank website. It contains the Human Development Indexes of every country. This was of great use because we wanted to see the connection between the spatial democracy index and happiness index of a country. We wanted to do this because this could help us with a greater connection between greater representation and the happiness/well-being of a country.


```{r}
final_df <- final_df %>%
  filter(is_democracy==TRUE)%>%
  filter(!pop_prop==Inf)%>%
  mutate(pop_prop = population/rowSums(cbind(upper_house_members,lower_house_members, third_house_members), na.rm=TRUE))
```

```{r}
final_df %>%
  select(country, year, population, pop_prop) %>%
  slice_max(order_by = pop_prop, n=15) %>%
  mutate(
    population = scales::comma(population),
    pop_prop = scales::comma(round(pop_prop, 0))
  ) %>%
  knitr::kable(
    col.names = c(
      "Country",
      "Year",
      "Population",
      "People per Representative"
    ),
    caption = "Population-adjusted representation for selected democratic countries.",
    align = "c"
  )
```

This table illustrates how population and representation levels change over time for different countries. This can highlight changes in historical and geopolitical trends, such as the rise of democratic nations over the past seventy years, and the challenges of some countries with overcoming monarchic structures. Since we are measuring in people per representative, this selection shows the least representative countries adjusted per population. Overall, we can see evolution of a country's structure by the progression of their representation numbers every five years.


```{r}
avg_dem <- function(country) {
  dem_df %>%
    filter(country_name == country) %>%
    summarize(avg = mean(spatial_democracy, na.rm = TRUE)) %>%
    pull(avg)
}

avg_df <- dem_df %>%
  distinct(country_name) %>%
  mutate(Average_Democracy = map_dbl(country_name, avg_dem))


```

```{r}
avg_df %>%
  mutate(
    Average_Democracy = round(Average_Democracy, 3)
  ) %>%
  slice_max(order_by = Average_Democracy, n=15)%>%
  knitr::kable(
    col.names = c(
      "Country",
      "Average Democracy Score"
    ),
    caption = "Average spatial democracy score by country.",
    align = "c"
  )
```

This table summarizes the average democracy score for each country. It provides a simplified measure of long-term democratic performance across nations. We do recognize however, that average does not provide a comprehensive look into the country's quality of democracy. For example, countries like Japan, South Korea, and former USSR countries had quick turnaround to democratic structures. 


```{r}
# Identifying which countries to label:
top_wb  <- hdiXdem_df %>% slice_max(avg_well_being, n = 3, with_ties = FALSE)
low_wb  <- hdiXdem_df %>% slice_min(avg_well_being, n = 3, with_ties = FALSE)
top_dem <- hdiXdem_df %>% slice_max(avg_democracy, n = 2, with_ties = FALSE)
low_dem <- hdiXdem_df %>% slice_min(avg_democracy, n = 2, with_ties = FALSE)

label_countries <- bind_rows(top_wb, low_wb, top_dem, low_dem) %>%
  distinct(country_name, .keep_all = TRUE)
```

```{r}
scatter1 <- ggplot(
  hdiXdem_df,
  aes(x = avg_democracy, y = avg_well_being)
) +
  geom_point(size = 2.8, alpha = 0.85, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1, color = "orange") +
  geom_text_repel(
    data = label_countries,
    aes(label = country_name),
    size = 3,
    max.overlaps = 20
  ) +
  scale_x_continuous(labels = scales::number_format(accuracy = 0.1)) +
  scale_y_continuous(labels = scales::number_format(accuracy = 0.1)) +
  labs(
    title = "Democracy and Well-Being Are Positively Associated",
    subtitle = "Country averages from 2011 to 2020",
    x = "Average Spatial Democracy Score",
    y = "Average Well-Being Score"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold")
  )

```

```{r}
#| fig-alt: Scatter plot of average spatial democracy scores versus average well-being for countries between 2011 and 2020. Each point represents a country. The points form an upward pattern, indicating that countries with higher democracy scores tend to have higher average well-being. A few countries at the high end are Finland, Switzerland, and Australia, and a few countries at the low end are Burundi, Central African Republic, and South Sudan.
#| warning: false
scatter1
```

This scatter plot shows a clear positive relationship between democracy and well-being. This is a very important visualization because we can obtain a preliminary sense of an answer to our question, that yes, a good quality democracy does tend to bolster the happiness and well-being of its constituents. Therefore, if this does prove to be statistically true, then the next important task becomes finding ways to improve the quality of a democracy to foster well-being.


```{r}
#| fig-cap: "Distribution of well-being across different levels of democracy."

hdiXdem_df %>%
  mutate(
    democracy_group = case_when(
      avg_democracy < 0.33 ~ "Low",
      avg_democracy < 0.66 ~ "Medium",
      TRUE ~ "High"),
    democracy_group = factor(democracy_group, levels = c("Low", "Medium", "High"))
  ) %>%
  ggplot(aes(x = democracy_group, y = avg_well_being, fill = democracy_group)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Well-Being Increases with Democracy Level",
    x = "Democracy Level",
    y = "Average Well-Being"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

This visualization shows that countries with higher levels of democracy generally have higher and more consistent well-being scores. The upward shift in medians reinforces the positive association. This is largely just a different angle of the same conclusion that a high quality democracy is associated with a higher average well-being. Again, this gives us further reason to conduct a statistical test to determine whether or not democracy level is truly associated with well being.


```{r}
final_df <- final_df %>%
  filter(!pop_prop==Inf)
```


```{r}
#| tbl-cap: "Linear model predicting average well-being from average spatial democracy."

rq2_model <- lm(avg_well_being ~ avg_democracy, data = hdiXdem_df)

tidy(rq2_model) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = pvalue(p.value)
  ) %>%
  rename(
    Term = term,
    Estimate = estimate,
    `Standard Error` = std.error,
    `t-value` = statistic,
    `p-value` = p.value
  ) %>%
  knitr::kable(
    caption = "Linear regression model predicting average well-being from average spatial democracy."
  )
```


The linear regression results indicate a strong, statistically significant positive relationship between average spatial democracy and average well-being. The coefficient for average democracy is 1.557, meaning that for every one-unit increase in the democracy score (on a 0–1 scale), the model predicts an increase of approximately 1.56 points in a country’s average well-being score.

The p-value for this coefficient is < 0.001, providing strong evidence that this relationship is statistically significant at the 0.05 level. This suggests that the observed positive association is very unlikely to be due to random chance.

The intercept of 4.547 represents the predicted well-being score when the democracy index is zero. This value is useful in practice, as some countries we are investigating have democracy indices of zero, so it serves as a baseline for the model.

Overall, these results support the conclusion that countries with higher levels of democracy tend to exhibit higher levels of well-being, reinforcing the idea that democratic structures may be associated with improved quality of life.

However, this model does not imply causation, and other factors such as economic development or regional differences may also influence well-being.
